Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array

### 2024-01-23 docs/manuscript/all_internal_comments/internal_comments_5/difference_file/new.tex

The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ can  then be calculated at each timestep from the Kalman filter output according to
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ .
\end{eqnarray}
The total log-likelihood for the entire sequence is
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
\end{eqnarray}
Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e. $\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$. Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$. If $\boldsymbol{\hat{\theta}}$ is close to the true underlying parameters $\boldsymbol{\theta}$, then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are minimized and $\mathcal{L}$ is maximised. This is illustrated with synthetic data in Figure \ref{fig:kalman_example}. In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations \eqref{eq:spinevol}-- \eqref{eq:xieqn}, \eqref{eq:measurement}, and \eqref{eq:g_func} for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$. The Kalman filter recovers the evolution of $f_{\rm p}^{(1)}(t)$ with high fidelity; the estimate of $\hat{f}_{\rm p}^{(1)}(t)$ (left top panel, blue curve) overlaps almost perfectly with the true $f_{\rm p}^{(1)}(t)$ (left top panel, green curve). The predicted state $\hat{f}_{\rm p}^{(1)}(t)$ is converted  into a predicted measurement $\hat{f}_{\rm m}^{(1)}(t)$ (middle panel, magenta curve), which again overlaps almost perfectly with the true measurement. The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel). By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$) to the Kalman filter, where $\Omega$ is displaced from its true value by $20 \%$. In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates. It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel).

%n Figure ; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity. The residuals correspond to random noise and are normally distributed. Conversely, if $\boldsymbol{\hat{\theta}}$ is not close to the true parameters then the filter is unable to recover the state evolution. This is demonstrated in Figure \ref{fig:25MB_bfs} where the Kalman filter was run with $\boldsymbol{\hat{\theta}}$ slightly perturbed away from the true values. In this case the filter cannot track the state variable accurately and the residuals are no longer Gaussian.

\subsection{Nested sampling}\label{sec:nested_sampling}
We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
 We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper. Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}. It also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies. Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. \newline

@@

As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
\begin{eqnarray}
	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
\end{eqnarray}
and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach. We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$. In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}. We see that the synthetic and empirical residuals are qualitatively similar. For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ for PSR J1939$+$2134.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J1939+2134. The actual residuals are sourced from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$. The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}. We set the TOA uncertainty to be a constant, $0.1 \mu$s. The blue and orange residuals are qualitatively similar by inspection. Similar results are obtained for the other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} by tuning the value of $\sigma^{(n)}$.}
	\label{fig:qualitative_compare}
\end{figure}
\subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
We generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
\begin{enumerate}[leftmargin=2em]
	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}.
\end{enumerate}
Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \ref{sec:synt_pta}. For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence. \newline

The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, through
\begin{equation}
	\sigma_{\rm m} \approx \sigma_{\rm TOA} f_{\rm p}^{(n)}  \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
\end{equation}
Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct. Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise. For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s, Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz. The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz. Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred for the sake of simplicity; this assumption can be relaxed easily when analysing real data. Whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}(t)$ is constructed from a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar. \newline

Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} \Delta B_t | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta B_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}). To fix the problem, we subtract the deterministic frequency evolution and track the new variable
\begin{equation}
	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ , \label{eq:hetero1}
\end{equation}  
equivalent to a change of variables. We similarly modify the measurement variable to be
\begin{equation}
	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ , \label{eq:hetero2}
\end{equation}
where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice. For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately. Equation \eqref{eq:measurement} is then updated to read 
\begin{equation}
	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
	\label{eq:measurement_cov}
\end{equation}
We emphasise that the change of variables in Equations \eqref{eq:hetero1} and \eqref{eq:hetero2} is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g. long double, quadruple). It does not remove any degrees of freedom nor does it involve an approximation. In particular $f_{\rm em}^{(n)}(t_1)$
and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.

\section{Representative end-to-end analysis} \label{sec:rep_example}
In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source. The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper. The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space}--\ref{sec:discussion}. To assist with reproducibility, we present intermediate outputs from each step. In Section \ref{sec:priors} we define the priors chosen on $\boldsymbol{\theta}$.

In Section \ref{sec:earth_psr_terms} we do this...

\textcolor{red}{TK to do}

In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise. In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates. In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$. \newline

The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative. The injected elements of $\boldsymbol{\theta}_{\rm gw}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}. The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$, described in Section \ref{sec:synt_pta}, are also recorded in Table \ref{tab:parameters_and_priors}.

@@

The parameters that govern the deterministic evolution of $f_{\rm p}^{(n)}(t)$, namely $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, are well-determined by radio timing observations. We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues \citep{Manchester2005}. For the 12.5 year NANOGrav pulsars the median fractional errors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$ are $\pm 2.68 \times 10^{-13} \%$ and $\pm 2.31 \times 10^{-3} \%$ respectively. In this paper we adopt uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database \citep{Manchester2005}. Adopting more relaxed priors tests the method stringently. The results below confirm that the method estimates ${\boldsymbol{\theta}}_{\rm gw}$ accurately, whether the priors are relaxed or narrow. \newline

>>>>

The Gaussian log-likelihood of obtaining ${\boldsymbol{Y}}_i$ given ${\boldsymbol{\hat X}}_i$ can  then be calculated at each timestep from the Kalman filter output according to
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ .
\end{eqnarray}
The total log-likelihood for the entire sequence is
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{M} \log \mathcal{L}_i \ . \label{eq:likelihood}
\end{eqnarray}
Given ${\boldsymbol{Y}}_1, \dots, {\boldsymbol{Y}}_M$, $\mathcal{L}$ is a function of the estimates ${\boldsymbol{\hat \theta}}$ of the static parameters passed to the Kalman filter, i.e. $\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\hat \theta})$. Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of $\boldsymbol{\hat \theta}$. If $\boldsymbol{\hat{\theta}}$ is close to the true underlying parameters $\boldsymbol{\theta}$, then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are minimized and $\mathcal{L}$ is maximised. This is illustrated with synthetic data in Figure \ref{fig:kalman_example}. In the left column, a time series of $f_{\rm m}^{(1)}(t)$ including Gaussian noise (middle panel, red curve) is generated from Equations \eqref{eq:spinevol}-- \eqref{eq:xieqn}, \eqref{eq:measurement}, and \eqref{eq:g_func} for a single pulsar and fed into the Kalman filter along with the true static parameters ${\boldsymbol{\hat \theta}} = {\boldsymbol{\theta}}$. The Kalman filter recovers the evolution of $f_{\rm p}^{(1)}(t)$ with high fidelity; the estimate of $\hat{f}_{\rm p}^{(1)}(t)$ (left top panel, blue curve) overlaps almost perfectly with the true $f_{\rm p}^{(1)}(t)$ (left top panel, green curve). The predicted state $\hat{f}_{\rm p}^{(1)}(t)$ is converted  into a predicted measurement $\hat{f}_{\rm m}^{(1)}(t)$ (middle panel, magenta curve), which again overlaps almost perfectly with the true measurement. The residuals $\epsilon(t)$ between the true and predicted measurements are small ($\lesssim 0.1\%$) and normally distributed (left bottom panel). By contrast, in the right column, the exercise is repeated while passing slightly incorrect static parameters (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$) to the Kalman filter, where $\Omega$ is displaced from its true value by $20 \%$. In this case the Kalman filter fails to track $f_{\rm p}^{(1)}(t)$ accurately, as the discrepancy between the blue and green curves in the top panel of the right-hand column indicates. It similarly fails to predict $f_{\rm m}^{(1)}(t)$ accurately, as shown by the discrepancy between the red and magenta curves in the middle panel, and the residuals are no longer distributed normally (right bottom panel). In Section \ref{sec:nested_sampling}, we explain how to combine the Kalman filter with a nested sampler to iteratively guide ${\boldsymbol{\hat \theta}}$ towards the true value of ${\boldsymbol{\theta}}$.

\subsection{Nested sampling}\label{sec:nested_sampling}
We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
 We estimate the posterior distribution and the model evidence through nested sampling \citep{Skilling} in this paper. Nested sampling evaluates marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}. It also approximates the posterior by returning samples from $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does so by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is its ability to compute $\mathcal{Z}$, on which model selection relies. Nested sampling is also computationally efficient and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, neutron star astrophysics \citep{Myers2021MNRAS.502.3113M,Meyers2021,Melatos2023}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For reviews of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. In gravitational wave research it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. \newline

@@

As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
\begin{eqnarray}
	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
\end{eqnarray}
and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach. We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$. In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}. We see that the synthetic and empirical residuals are qualitatively similar. For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is much less than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ for PSR J1939$+$2134.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J1939+2134. The actual residuals are sourced from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$. The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}. We set the TOA uncertainty to be a constant, viz. $0.1 \mu$s. The blue and orange residuals are qualitatively similar by inspection. Similar results are obtained for the other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} by tuning the value of $\sigma^{(n)}$.}
	\label{fig:qualitative_compare}
\end{figure}
\subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
We generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
\begin{enumerate}[leftmargin=2em]
	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}.
\end{enumerate}
Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \ref{sec:synt_pta}. For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence. \newline

The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, through
\begin{equation}
	\sigma_{\rm m} \approx \sigma_{\rm TOA} f_{\rm p}^{(n)}  \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
\end{equation}
Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct. Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise. For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 10^2$ Hz observed with weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s, Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz. The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz. Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred for the sake of simplicity; this assumption can be relaxed easily when analysing real data. Whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}(t)$ is constructed from a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar. \newline

Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} dB(t) | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta B_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}). To fix the problem, we subtract the deterministic frequency evolution and track the new variable
\begin{equation}
	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ , \label{eq:hetero1}
\end{equation}  
equivalent to a change of variables. We similarly modify the measurement variable to be
\begin{equation}
	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ , \label{eq:hetero2}
\end{equation}
where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic evolution based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice. For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately. Equation \eqref{eq:measurement} is then updated to read 
\begin{equation}
	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
	\label{eq:measurement_cov}
\end{equation}
We emphasise that the change of variables in Equations \eqref{eq:hetero1} and \eqref{eq:hetero2} is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g. long double, quadruple). It does not remove any degrees of freedom nor does it involve an approximation. In particular $f_{\rm em}^{(n)}(t_1)$
and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, i.e. Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.

\section{Representative end-to-end analysis} \label{sec:rep_example}
In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source. The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper. The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space}--\ref{sec:discussion}. To assist with reproducibility, we present intermediate outputs from each step. In Section \ref{sec:priors} we define the priors chosen on $\boldsymbol{\theta}$. In Section \ref{sec:earth_psr_terms} we define the measurement equation used in the inference model by the Kalman filter. In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise. In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates. In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$. \newline

The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis are selected to be astrophysically reasonable and representative. The injected components of $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ are summarised in the second column of Table \ref{tab:parameters_and_priors}.

@@

The parameters that govern the deterministic evolution of $f_{\rm p}^{(n)}(t)$, namely $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$, are well-determined by radio timing observations. We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues \citep{Manchester2005}. For the 12.5 year NANOGrav pulsars the median fractional errors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$ are $\pm 2.68 \times 10^{-13} $ per cent and $\pm 2.31 \times 10^{-3}$ per cent respectively. In this paper we adopt uniform priors on $f_{\rm em}^{(n)}(t_1)$ and $\dot{f}_{\rm em}^{(n)}(t_1)$, which extend $\pm 10^3 \eta_f^{(n)}$ and $\pm 10^3 \eta_{\dot{f}}^{(n)}$ respectively about the central, injected values, where $\eta_f^{(n)}$ and $\eta_{\dot{f}}^{(n)}$ denote the errors quoted in the ATNF Pulsar Database \citep{Manchester2005}. Wider-than-necessary priors test the method more stringently than narrow priors. The results below confirm that the method estimates ${\boldsymbol{\theta}}_{\rm gw}$ accurately, whether the priors are wide or narrow. \newline
